---
title: "Graph 2 - Eda Gokdogan"
output: html_document
date: "2024-05-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())  
library(ggplot2)
library(dplyr)

crime_data <- read.csv("/Users/edagokdogan/Downloads/Crime_Data 2.csv")
```

```{r, echo=FALSE}
# formatting the function and extracting the month in order to categorize them as seasons 
crime_data$Occurred.Date <- as.character(crime_data$Occurred.Date)
crime_data$Date <- as.Date(crime_data$Occurred.Date, tryFormats = c("%m/%d/%Y"))
crime_data$Month <- format(crime_data$Date, "%m") #extracting the month so that we can categorize it by seasons
crime_data <- na.omit(crime_data)
#creating seasons column
seasons <- function(month) {
  month <- as.numeric(month)  # Convert month to numeric inside the function
  if (month %in% c(12, 01, 02)) {
    return("Winter")
  } else if (month %in% c(03, 04, 05)) {
    return("Spring")
  } else if (month %in% c(06, 07, 08)) {
    return("Summer")
  } else {
    return("Fall")
  }
}


plot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)
  plots <- c(list(...), plotlist)

  numPlots = length(plots)
  if (is.null(layout)) {
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    for (i in 1:numPlots) {
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}

#the graph is not working # combining all and creating one graph  
crime_data$Season <- sapply(crime_data$Month, seasons)

p <- ggplot(crime_data, aes(x = Season, fill = Crime.Subcategory, na.rm=TRUE)) +
  geom_bar(stat = "count", color= "black") +
  ggtitle(label = "Histogram of Crime Rates by Seasons and Types of Crimes") + 
  xlab("Seasons") +
  ylab("Count of Crimes") +
  theme_minimal() + 
 theme(
   legend.position = "right",
    plot.title = element_text(size = 13, face = "bold"),
    axis.title.x = element_text(size =10, face="bold"),
    axis.title.y = element_text(size = 10, face="bold"),
    axis.text.x = element_text(size = 7),
    axis.text.y = element_text(size = 7),
    strip.text.y = element_text(size = 3, face = "bold"),
    legend.title = element_text(size = 6, face = "bold"),
    legend.text = element_text(size = 5)
 )
print(p)


```

 
  
I chose to create this figure because this figure not only shows the major crime subcategories of crimes occurred in the Seattle between 1973-2018, but also provides information about what types of crimes happened in which seasons. This figure reveals information about the highest category of the crime in Seattle  and reveals that the highest crime happened in car prowls in all seasons. Furthermore, the graph provides information about which season highest crime occurred in Seattle. According to the graph, spring was the highest season that the crime has occurred.
